---
title: "EXERCISE 11.2"
author: "Ramsey King"
date: '2021-05-29'
output:
  html_document: default
  pdf_document: default
  word_document: default
# bibliography: bibliography.bib
---

```{r setup, echo=FALSE}
#knitr::opts_knit$set(root.dir = 'C:/Users/Ramsey/Documents/GitHub/dsc520/')
knitr::opts_knit$set(root.dir = '/home/ramsey/GitHub/dsc520')
```

## In this problem, you will use the nearest neighbors algorithm to fit a model on two simplified datasets. 

## The first dataset (found in binary-classifier-data.csv) contains three variables; label, x, and y. The label variable is either 0 or 1 and is the output we want to predict using the x and y variables (You worked with this dataset last week!). 

## The second dataset (found in trinary-classifier-data.csv) is similar to the first dataset except that the label variable can be 0, 1, or 2.

### 1. Plot the data from each dataset using a scatter plot.

```{r scatter_plot_binary, echo=FALSE}
## Read the csv file in the binary_classifier variable
binary_data <- read.csv("assignments/week11/binary-classifier-data.csv")

## load the ggplot2 library
library(ggplot2)

## Scatter plot of the binary data set.
ggplot(binary_data, aes(x = x, y = y)) +
  geom_point()
```
```{r scatter_plot_trinary, echo=FALSE}
## Read the csv file in the binary_classifier variable
trinary_data <- read.csv("assignments/week11/trinary-classifier-data.csv")

## load the ggplot2 library
library(ggplot2)

## Scatter plot of the tridary data set.
ggplot(trinary_data, aes(x = x, y = y)) +
  geom_point()
```


### 2. Fit a k nearest neighbors’ model for each dataset for k=3, k=5, k=10, k=15, k=20, and k=25. Compute the accuracy of the resulting models for each value of k. Plot the results in a graph where the x-axis is the different values of k and the y-axis is the accuracy of the model.

```{r k_means_binary, echo=FALSE}
binary_train <- binary_data[, which(names(binary_data) != "label")]

set.seed(36912)
binaryk3 <- kmeans(x=binary_train, centers=3)
binaryk5 <- kmeans(x=binary_train, centers=5)
binaryk10 <- kmeans(x=binary_train, centers=10)
binaryk15 <- kmeans(x=binary_train, centers=15)
binaryk20 <- kmeans(x=binary_train, centers=20)
binaryk25 <- kmeans(x=binary_train, centers=25)

binarytrainIndex <- createDataPartition(binary_data$label, p = 0.7,
                                        list = F,
                                        times =1)
binary_train_dataset <- binary_data[binarytrainIndex,]
binary_test_dataset <- binary_data[-binarytrainIndex,]





```